The role of advection versus diffusion in intraventricular transport

### 2025-03-18 paper/main.tex
abstract draft

Please insert your abstract here

@@

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. Produced mostly by the choroid
plexus and partly by brain interstitial fluid,
CSF is a clear liquid that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013, Brodal2016}. By transporting growth factors, nutrients, neurotransmitters
and metabolites as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of the CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment~\cite{duy2022rethinking, kahle2024paediatric, wallmeier2022role},
but for which there is little agreement on optimal treatments~\cite{flannery2014pediatric}. 
Of particular importance is the fact that
hydrocephalus is observed in animal models and humans with deficient motile
cilia~\cite{Afzelius2004Cilia, Olstad2019CiliaryDevelopment, Mitchison2017Motile,
Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain functioning.

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movement~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Motile cilia are hair-like structures that line the ventricular surface of the brain,
known as the ependyma~\cite{mitchell2007evolution, Ringers2020Role, spassky2013motile}. 
The cilia beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental endeavors are challenging to pursue due to
technical limitations: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia, Sawamoto2006NewBrain, ibanez2003beat};
however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF
flow and the accompanying transport. Similarly, while MRI technology
can measure CSF movement in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. Consequently, we lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport, and whether cilia
dysfunction alters CSF composition and thus leads to brain disorders.

>>>>

Cerebrospinal fluid (CSF) flow is integral to brain function.
The underlying mechanisms of cilia-mediated flow and transport
in the brain ventricles remain to be fully explored.
In this paper, we analyze and evaluate the role of motile cilia
in intraventricular CSF flow and transport. We combine computational
modeling with in vivo experiments that monitor transport of a photoconvertible
protein secreted in the brain ventricles.
We use the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF motion velocities and frequencies,
to develop a finite element model.
Our results show that while cilia contribute to advection of large particles,
diffusion plays a significant role in the transport of small molecules such as proteins.
We also found that cilia location and the geometry of the ventricular system impact solute distribution.
Altogether, our work offers both a computational framework that can be applied to other
ventricular system and new concepts of how molecules are transported within the brain and its ventricles.

@@

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. Produced mostly by the choroid
plexus and partly by brain interstitial fluid,
CSF is a clear liquid that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013, Brodal2016}. By transporting growth factors, nutrients, neurotransmitters
and metabolites, as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of the CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment~\cite{duy2022rethinking, kahle2024paediatric, wallmeier2022role},
but for which there is little agreement on optimal treatments~\cite{flannery2014pediatric}. 
Of particular importance is the fact that
hydrocephalus is observed in animal models and humans with deficient motile
cilia~\cite{Afzelius2004Cilia-relatedDiseases, Olstad2019CiliaryDevelopment, Mitchison2017Motile,
Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movement~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Motile cilia are hair-like structures that line the ventricular surface of the brain,
known as the ependyma~\cite{mitchell2007evolution, Ringers2020Role, spassky2013motile}. 
The cilia beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental endeavors are challenging to pursue due to
technical limitations: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF
flow and the accompanying transport. Similarly, while MRI technology
can measure CSF movement in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. Consequently, we lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport, and whether cilia
dysfunction alters CSF composition and thus leads to brain disorders.
